---
title: "Cardiopulmonary Bypass (CPB) Response Analysis in Tetralogy of Fallot (ToF) and Atrial Septal Defect (ASD)"
author: Sara Tahir
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Overview
This pipeline analyzes the transcriptomic response to cardiopulmonary bypass (CPB) in pediatric patients with Tetralogy of Fallot (ToF) and Atrial Septal Defect (ASD) stratified by sex.

**Platform:** Affymetrix Human Genome U133 Plus PM Array (GSE132176)

**Design:** Paired samples (Pre-CPB and Post-CPB) from the same individuals.

## Setup and Read data
```{r package-setup, message = FALSE}
# Load/Install Manager
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse, gplots, here,      # CRAN
  GEOquery, oligo, limma, edgeR, hthgu133pluspm.db # Bioconductor
)
```

```{r read-data, message = FALSE}
# Read in geo dataset
geo_data <- getGEO(filename = "GSE132176_series_matrix.txt.gz")
metadata <- pData(geo_data)

cel_files <- list.celfiles(path = "GSE132176_RAW", full.names = TRUE)
file_gsms <- stringr::str_extract(cel_files, "GSM[0-9]+")
ordered_files <- cel_files[match(metadata$geo_accession, file_gsms)]
raw_data <- read.celfiles(ordered_files)
```

## Pre-Processing and Normalization
Use the **Robust Multi-array Average (RMA)** algorithm to perform background subtraction, quantile normalization, and log2 summarization.
*(Note: Unable to filter based on PA calls (Presence/Absence) because "this is a PM-only array, so MAS5-style P/A calls are not applicable."*
```{r normalization}
# Pre-RMA Distribution Check
boxplot(raw_data, target = "core", main = "Pre-Normalization: Raw Intensities", las = 2, cex.axis = 0.5)
# Use RMA to do background subtraction, normalization, and summarization
eset <- rma(raw_data)
# Get ExpressionFeatureSet
expression_matrix <- exprs(eset)
# Post-RMA Distribution Check
boxplot(expression_matrix, main = "Post-RMA Normalization: Log2 Intensities", las = 2, cex.axis = 0.5)
```

## Intensity Filtering
To increase power, drop probes that are mostly just background noise. A density plot of the probe means shows the noise floor is around an instensity of log(10), 3.31

```{r instensity-filtering}
# Average expression for every probe across all samples
probe_means <- rowMeans(expression_matrix)
# Density distribution of averages
plot(density(probe_means), 
     main = "Density of Probe Average Intensities", 
     xlab = "Log2 Intensity (RMA Normalized)")
# Red line at the log2(10) threshold (3.32)
abline(v = log2(10), col = "red", lwd = 2, lty = 2)
keep_intensity <- rowMeans(expression_matrix) > log2(10)
expression_matrix_filtered <- expression_matrix[keep_intensity, ]
```

## Annotation of Genes
``` {r annotate-genes}
# Label genes
gene_symbols <- mapIds(hthgu133pluspm.db,
                      keys = rownames(expression_matrix_filtered),
                      column = "SYMBOL",
                      keytype = "PROBEID",
                      multiVals = "first")

# Filter out probes that don't map to any known gene (NAs)
valid_probes <- !is.na(gene_symbols)
expression_matrix_filtered <- expression_matrix_filtered[valid_probes, ]
hgnc_symbols_filtered <- gene_symbols[valid_probes]
```

## Transcriptomic Sex Identification
Metadata in this dataset lacks male/female designation, so to distinguish between males and females the expression of the following genes was analyzed.

Female only: XIST

Male only: KDM5D, RPS4Y1, NLGN4Y, EIF1AY
```{r sex-determination}
# Create a temporary collapsed matrix just for the sex calculation math
temp_annotated <- avereps(expression_matrix_filtered, ID = hgnc_symbols_filtered)

sex_genes <- c("XIST", "KDM5D", "RPS4Y1", "EIF1AY", "NLGN4Y") 
available_sex_genes <- sex_genes[sex_genes %in% rownames(temp_annotated)]

# Subset matrix to contain just sex genes and transpose so that samples are rows
sex_matrix <- t(temp_annotated[available_sex_genes, ])

# Heiarchical clustering
dist_sex <- dist(sex_matrix)
hclust_sex <- hclust(dist_sex)

plot(hclust_sex, main="Dendrogram: Sex Determination", xlab="Samples", cex=0.5)

heatmap.2(t(sex_matrix), 
          trace = "none", scale = "row",
          col = colorRampPalette(c("blue", "white", "red"))(50),
          main = "Sex Gene Expression Profile")

clusters <- cutree(hclust_sex, k = 2)

# Calculate the mean XIST expression for both clusters
xist_c1 <- mean(sex_matrix[clusters == 1, "XIST"])
xist_c2 <- mean(sex_matrix[clusters == 2, "XIST"])

# The cluster with the higher XIST expression is Female
female_cluster <- ifelse(xist_c1 > xist_c2, 1, 2)

# Build the final sex dataframe for the pipeline
sexDf <- data.frame(
  sampleID = rownames(sex_matrix),
  Cluster = clusters
) %>%
  mutate(
    # Clean up the sample IDs to match metadata
    sampleID = gsub("^(GSM[0-9]+)_.*$", "\\1", basename(sampleID)),
    # Assign the text labels
    transcriptomicSex = ifelse(Cluster == female_cluster, "Female", "Male")
  )
```

## Metadata formatting
```{r metadata-formatting}
metadata$sampleID <- rownames(metadata)

# Merge sex and create clean factors
metadata <- metadata %>%
  left_join(sexDf %>% dplyr::select(sampleID, transcriptomicSex), by = c("geo_accession" = "sampleID")) %>%
  mutate(
    Time = factor(ifelse(grepl("before", `time:ch1`), "Pre", "Post"), levels = c("Pre", "Post")),
    Disease = factor(`disease state:ch1`),
    PatientID = factor(gsub("patient ", "Pt", `individual:ch1`)), 
    Sex = factor(transcriptomicSex)
  )

# Apply clean names (e.g., "TOF_Pre_Male_Pt1")
clean_sample_names <- paste(metadata$Disease, metadata$Time, metadata$Sex, metadata$PatientID, sep = "_")
colnames(expression_matrix_filtered) <- clean_sample_names
rownames(metadata) <- clean_sample_names
```

## Sample Quality Control
Used PCA plots and hierarchical clustering. 
*(Note: Unable to unable to create NUSE/RLE plots because R crashes upon calling fitProbeLevelModel)"*U

``` {r sample-qc}
# Hierarchical clustering
dist_mat <- dist(t(expression_matrix_filtered))
hclust_res <- hclust(dist_mat)
plot(hclust_res, 
     main = "Hierarchical Clustering of Samples", 
     xlab = "Samples", 
     sub = "", 
     cex = 0.6) # cex shrinks the text so patient IDs don't overlap as badly

# Heatmap of top variable genes
# Calculate variance for all genes across all samples
gene_vars <- apply(expression_matrix_filtered, 1, var)
# Grab top 500 genes
top_var_genes <- names(sort(gene_vars, decreasing = TRUE))[1:500]

heatmap.2(expression_matrix_filtered[top_var_genes, ], 
          trace = "none",         
          scale = "row",          # Scales the colors per gene (row) so extremes don't wash out the plot
          col = colorRampPalette(c("blue", "white", "red"))(50),
          main = "Top 500 Variable Genes",
          cexCol = 0.6,           
          margins = c(10, 5)) 

# PCA Plot
pca_res <- prcomp(t(expression_matrix_filtered), scale. = TRUE)
pca_df <- as.data.frame(pca_res$x)
pca_plot_data <- cbind(pca_df, metadata)

pca_res.var <- pca_res$sdev^2
pca_res.var.per <- round(pca_res.var/sum(pca_res.var)*100,1)
barplot(pca_res.var.per, main = "Scree Plot", xlab = "Principle Component", ylab = "Percent Variation")

# Create a combined grouping variable for the colors
pca_plot_data$Disease_Sex <- paste(pca_plot_data$Disease, pca_plot_data$Sex, sep = "_")

ggplot(pca_plot_data, aes(x = PC1, y = PC2, color = Disease_Sex, shape = Time)) +
  geom_point(size = 4, alpha = 0.8) +
  theme_bw() +
  
  # Assign complementary light/dark shades so the diseases group visually
  scale_color_manual(
    name = "Disease & Sex",
    values = c(
      "ASD_Female" = "orange", 
      "ASD_Male" = "firebrick", 
      "TOF_Female" = "deepskyblue", 
      "TOF_Male" = "darkblue"
    )
  ) +
  
  labs(
    title = "PCA of Filtered Expression Data",
    subtitle = "Single Panel: Disease/Sex (Color) and Time (Shape)",
    x = paste0("PC1 (", round(summary(pca_res)$importance[2,1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_res)$importance[2,2] * 100, 1), "%)")
  )
```

## Batch Effect Assessment
Physical scan dates were extracted to assess temporal batch effects.
```{r batch-effect-check}
raw_dates <- protocolData(raw_data)$dates

metadata$ScanDate <- as.factor(as.Date(raw_dates, format = "%Y/%m/%d"))
metadata$ScanYear <- as.factor(format(as.Date(raw_dates, format = "%Y/%m/%d"), "%Y"))

metadata$SampleGroup <- paste(metadata$Disease, metadata$Time, sep = "_")
table(Group = metadata$SampleGroup, Date = metadata$ScanDate)

pca_plot_data$ScanDate <- metadata$ScanDate
pca_plot_data$ScanYear <- metadata$ScanYear

ggplot(pca_plot_data, aes(x = PC1, y = PC2, color = ScanDate, shape = Time)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_bw() +
  labs(
    title = "PCA Checking for Batch Effects (Scan Date)",
    subtitle = "Shape denotes Pre (Circle) vs Post (Triangle)",
    x = paste0("PC1 (", round(summary(pca_res)$importance[2,1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_res)$importance[2,2] * 100, 1), "%)")
  )

year_colors <- c("2013" = "dodgerblue", "2014" = "firebrick")

ggplot(pca_plot_data, aes(x = PC2, y = PC3, color = ScanYear, shape = Time)) +
  geom_point(size = 4, alpha = 0.8) +
  theme_bw() +
  scale_color_manual(values = year_colors) +
  labs(
    title = "PCA Checking for Batch Effects (Scan Year)",
    subtitle = "Visualizing the temporal confounding effect",
    x = paste0("PC2 (", round(summary(pca_res)$importance[2,2] * 100, 1), "%)"),
    y = paste0("PC3 (", round(summary(pca_res)$importance[2,3] * 100, 1), "%)")
  )

```

## Limma DGE
``` {r limma-model}
# Primary grouping factor, ex: TOF_Post_Male
metadata$group <- factor(paste(metadata$Disease, metadata$Time, metadata$Sex, sep = "_"))
metadata$ScanYear <- as.factor(metadata$ScanYear)

# Set up design matrix, including batch by scan year (2013 or 2014)
design <- model.matrix(~0 + group + ScanYear, data = metadata)
colnames(design) <- gsub("group", "", colnames(design))

# duplicateCorrelation estimates the internal consistency of each child's samples
corfit <- duplicateCorrelation(expression_matrix_filtered, design, block = metadata$PatientID)

fit <- lmFit(expression_matrix_filtered, design, block = metadata$PatientID, correlation = corfit$consensus)

# Create contrasts
contrastMatrix <- makeContrasts(
  TOF_Male_Response = TOF_Post_Male - TOF_Pre_Male,
  TOF_Female_Response = TOF_Post_Female - TOF_Pre_Female,
  ASD_Male_Response = ASD_Post_Male - ASD_Pre_Male,
  ASD_Female_Response = ASD_Post_Female - ASD_Pre_Female,
  levels = design
)

# Final Statistics
fit2 <- contrasts.fit(fit, contrastMatrix)
fit2 <- eBayes(fit2, robust = TRUE)

# View Summaries
resultsSummary <- decideTests(fit2)
print(summary(resultsSummary))
vennDiagram(resultsSummary, cex = c(1, 0.6, 1))
```

## Post-Analysis and Result Generation
```{r post-analysis}
hgnc_complete_set_df <- read_tsv("hgnc_complete_set.txt") %>%
  dplyr::select(HGNC_Symbol = symbol)

contrasts_list <- colnames(contrastMatrix)
for (contrast_name in contrasts_list) {
  res_df <- topTable(fit2, coef = contrast_name, number = Inf, adjust.method = "BH")
  
  # Add HGNC gene symbol
  res_df_with_hgnc <- res_df %>%
    as_tibble(rownames = "Original_Probe_ID") %>% 
    mutate(HGNC_Symbol = hgnc_symbols_filtered[match(Original_Probe_ID, names(hgnc_symbols_filtered))])
  
  # Remove rows where HGNC_Symbol is NA
  res_df_with_hgnc <- res_df_with_hgnc %>%
    filter(!is.na(HGNC_Symbol))
  
  # Perform inner_join with the hgnc_complete_set data
  joined_results <- res_df_with_hgnc %>%
    inner_join(hgnc_complete_set_df %>% dplyr::select(HGNC_Symbol), by = "HGNC_Symbol")
  
  final_results_table <- joined_results %>%
    dplyr::arrange(adj.P.Val) %>% 
    dplyr::distinct(HGNC_Symbol, .keep_all = TRUE)
  
  final_results_table <- final_results_table %>%
    dplyr::select(HGNC_Symbol, logFC, adj.P.Val) %>%
    dplyr::rename(P.value = adj.P.Val)
  
  final_results_table <- final_results_table[!is.na(final_results_table$P.value), ]
  
  file_name <- paste0("dge_results_", contrast_name, "_final.csv")
  write_csv(final_results_table, file_name)
  
  message(paste("Saved:", file_name))
}
```


