---
title: "Cardiopulmonary Bypass (CPB) Response Analysis in Tetralogy of Fallot (ToF) and Atrial Septal Defect (ASD)"
author: Sara Tahir
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Overview
This pipeline analyzes the transcriptomic response to cardiopulmonary bypass (CPB) in pediatric patients with Tetralogy of Fallot (ToF) and Atrial Septal Defect (ASD) stratified by sex.

**Platform:** Affymetrix Human Genome U133 Plus PM Array (GSE132176)

**Design:** Paired samples (Pre-CPB and Post-CPB) from the same individuals.

## Setup and Read data
```{r package-setup, message = FALSE}
# Load/Install Manager
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse, gplots, here, findbars,     # CRAN
  GEOquery, oligo, limma, edgeR, hthgu133pluspm.db, sva, variancePartition, BiocParallel # Bioconductor
)
```

```{r read-data, message = FALSE}
# Read in geo dataset
geo_data <- getGEO(filename = "GSE132176_series_matrix.txt.gz")
metadata <- pData(geo_data)

cel_files <- list.celfiles(path = "GSE132176_RAW", full.names = TRUE)
file_gsms <- stringr::str_extract(cel_files, "GSM[0-9]+")
ordered_files <- cel_files[match(metadata$geo_accession, file_gsms)]
raw_data <- read.celfiles(ordered_files)
```

## Pre-Processing and Normalization
Use the **Robust Multi-array Average (RMA)** algorithm to perform background subtraction, quantile normalization, and log2 summarization.
*(Note: Unable to filter based on PA calls (Presence/Absence) because "this is a PM-only array, so MAS5-style P/A calls are not applicable."*
```{r normalization}
# Pre-RMA Distribution Check
boxplot(raw_data, target = "core", main = "Pre-Normalization: Raw Intensities", las = 2, cex.axis = 0.5)
# Use RMA to do background subtraction, normalization, and summarization
eset <- rma(raw_data)
# Get ExpressionFeatureSet
expression_matrix <- exprs(eset)
# Post-RMA Distribution Check
boxplot(expression_matrix, main = "Post-RMA Normalization: Log2 Intensities", las = 2, cex.axis = 0.5)
```

## Intensity Filtering
To increase power, drop probes that are mostly just background noise. Filter by expression cannot be used because this is microarray data, so instead check to make sure a gene is expressed above the noise in at least n patients, n being the smallest group size

```{r instensity-filtering}
min_samples <- 4 # ASD males are the smallest group size

noise_threshold <- median(expression_matrix) # Can also use log(10) value
is_expressed <- expression_matrix > noise_threshold
keep_probes <- rowSums(is_expressed) >= min_samples

expression_matrix_filtered <- expression_matrix[keep_probes, ]

# Check how many probes were removed
total_probes <- nrow(expression_matrix)
kept_probes <- nrow(expression_matrix_filtered)
removed_probes <- total_probes - kept_probes

cat("Total starting probes:", total_probes, "\n")
cat("Probes kept:", kept_probes, paste0("(", round((kept_probes/total_probes)*100, 1), "%)"), "\n")
cat("Probes removed (Noise):", removed_probes, paste0("(", round((removed_probes/total_probes)*100, 1), "%)"), "\n")

# Create plot to view data before and after filtering
pre_filter_means <- rowMeans(expression_matrix)
post_filter_means <- rowMeans(expression_matrix_filtered)

plot(density(pre_filter_means), 
     main = "Effect of Filtering on Probe Density", 
     xlab = "Log2 Intensity (RMA Normalized)", 
     col = "black", 
     lwd = 2,
     ylim = c(0, max(density(pre_filter_means)$y) * 1.2))

lines(density(post_filter_means), col = "red", lwd = 2)

legend("topright", 
       legend = c("Pre-Filter (All Probes)", "Post-Filter (Kept Probes)"), 
       fill = c("black", "red"))

# Plot Mean vs Variance
pre_mean <- rowMeans(expression_matrix)
pre_sd <- apply(expression_matrix, 1, sd)

plot(pre_mean, pre_sd, 
     pch = 16, cex = 0.5, col = rgb(0,0,0,0.1), # Transparent black dots
     main = "Mean-Variance Trend", 
     xlab = "Average Log2 Expression", 
     ylab = "Standard Deviation")

# Highlight the kept probes in blue
points(pre_mean[keep_probes], pre_sd[keep_probes], 
       pch = 16, cex = 0.5, col = rgb(0,0,1,0.1))

legend("topright", legend=c("Discarded Probes", "Kept Probes"), 
       col=c("black", "blue"), pch=16)
```

## Annotation of Genes
``` {r annotate-genes}
# Label genes
gene_symbols <- mapIds(hthgu133pluspm.db,
                      keys = rownames(expression_matrix_filtered),
                      column = "SYMBOL",
                      keytype = "PROBEID",
                      multiVals = "first")

# Filter out probes that don't map to any known gene (NAs)
valid_probes <- !is.na(gene_symbols)
expression_matrix_filtered <- expression_matrix_filtered[valid_probes, ]
hgnc_symbols_filtered <- gene_symbols[valid_probes]
```

## Transcriptomic Sex Identification
Metadata in this dataset lacks male/female designation, so to distinguish between males and females the expression of the following genes was analyzed.

Female only: XIST

Male only: KDM5D, RPS4Y1, NLGN4Y, EIF1AY
```{r sex-determination}
# Create a temporary collapsed matrix just for the sex calculation math
temp_annotated <- avereps(expression_matrix_filtered, ID = hgnc_symbols_filtered)

sex_genes <- c("XIST", "KDM5D", "RPS4Y1", "EIF1AY", "NLGN4Y") 
available_sex_genes <- sex_genes[sex_genes %in% rownames(temp_annotated)]

# Subset matrix to contain just sex genes and transpose so that samples are rows
sex_matrix <- t(temp_annotated[available_sex_genes, ])

# Hierarchical clustering
dist_sex <- dist(sex_matrix)
hclust_sex <- hclust(dist_sex)

plot(hclust_sex, main="Dendrogram: Sex Determination", xlab="Samples", cex=0.5)

heatmap.2(t(sex_matrix), 
          trace = "none", scale = "row",
          col = colorRampPalette(c("blue", "white", "red"))(50),
          main = "Sex Gene Expression Profile")

clusters <- cutree(hclust_sex, k = 2)

# Calculate the mean XIST expression for both clusters
xist_c1 <- mean(sex_matrix[clusters == 1, "XIST"])
xist_c2 <- mean(sex_matrix[clusters == 2, "XIST"])

# The cluster with the higher XIST expression is Female
female_cluster <- ifelse(xist_c1 > xist_c2, 1, 2)

# Build the final sex dataframe for the pipeline
sexDf <- data.frame(
  sampleID = rownames(sex_matrix),
  Cluster = clusters
) %>%
  mutate(
    # Clean up the sample IDs to match metadata
    sampleID = gsub("^(GSM[0-9]+)_.*$", "\\1", basename(sampleID)),
    # Assign the text labels
    transcriptomicSex = ifelse(Cluster == female_cluster, "Female", "Male")
  )
```

## Metadata formatting
```{r metadata-formatting}
metadata$sampleID <- rownames(metadata)

# Merge sex and create clean factors
metadata <- metadata %>%
  left_join(sexDf %>% dplyr::select(sampleID, transcriptomicSex), by = c("geo_accession" = "sampleID")) %>%
  mutate(
    Time = factor(ifelse(grepl("before", `time:ch1`), "Pre", "Post"), levels = c("Pre", "Post")),
    Disease = factor(`disease state:ch1`),
    PatientID = factor(gsub("patient ", "Pt", `individual:ch1`)), 
    Sex = factor(transcriptomicSex)
  )

# Apply clean names (e.g., "TOF_Pre_Male_Pt1")
clean_sample_names <- paste(metadata$Disease, metadata$Time, metadata$Sex, metadata$PatientID, sep = "_")
colnames(expression_matrix_filtered) <- clean_sample_names
rownames(metadata) <- clean_sample_names

# Add scan dates and scan years to metadata
raw_dates <- protocolData(raw_data)$dates
metadata$ScanDate <- as.factor(as.Date(raw_dates, format = "%Y/%m/%d"))
metadata$ScanYear <- as.factor(format(as.Date(raw_dates, format = "%Y/%m/%d"), "%Y"))
metadata$SampleGroup <- paste(metadata$Disease, metadata$Time, sep = "_")
```

## Outlier Removal Toggle
```{r outlier-removal}
remove_outlier <- TRUE

if(remove_outlier) {
  # Add the specific samples you want to drop here.
  # Due to the paired design, it is recommended to drop both the Pre and Post samples for a patient.
  targets <- c("ASD_Pre_Female_Pt4", "ASD_Post_Female_Pt4")
  
  # Remove from expression matrix
  expression_matrix_filtered <- expression_matrix_filtered[, !(colnames(expression_matrix_filtered) %in% targets)]
  
  # Remove from metadata
  metadata <- metadata[!(rownames(metadata) %in% targets), ]
  
  message(paste("Outliers removed:", paste(targets, collapse = ", ")))
} else {
  message("No samples removed.")
}
```

## Sample Quality Control
Used hierarchical clustering and PCA plots
*(Note: Unable to unable to create NUSE/RLE plots because R crashes upon calling fitProbeLevelModel)"*U

``` {r sample-qc}
# Hierarchical clustering
dist_mat <- dist(t(expression_matrix_filtered))
hclust_res <- hclust(dist_mat)
plot(hclust_res, 
     main = "Hierarchical Clustering of Samples", 
     xlab = "Samples", 
     sub = "", 
     cex = 0.6) # cex shrinks the text so patient IDs don't overlap as badly

# Heatmap of top variable genes
# Calculate variance for all genes across all samples
gene_vars <- apply(expression_matrix_filtered, 1, var)
# Grab top 500 genes
top_var_genes <- names(sort(gene_vars, decreasing = TRUE))[1:500]

heatmap.2(expression_matrix_filtered[top_var_genes, ], 
          trace = "none",         
          scale = "row",          # Scales the colors per gene (row) so extremes don't wash out the plot
          col = colorRampPalette(c("blue", "white", "red"))(50),
          main = "Top 500 Variable Genes",
          cexCol = 0.6,           
          margins = c(10, 5)) 

# PCA Plot
pca_res <- prcomp(t(expression_matrix_filtered), scale. = TRUE)
pca_df <- as.data.frame(pca_res$x)
pca_plot_data <- cbind(pca_df, metadata)

pca_res.var <- pca_res$sdev^2
pca_res.var.per <- round(pca_res.var/sum(pca_res.var)*100,1)
barplot(pca_res.var.per, main = "Scree Plot", xlab = "Principle Component", ylab = "Percent Variation")

# Create a combined grouping variable for the colors
pca_plot_data$Disease_Sex <- paste(pca_plot_data$Disease, pca_plot_data$Sex, sep = "_")

ggplot(pca_plot_data, aes(x = PC1, y = PC2, color = Disease_Sex, shape = Time)) +
  geom_point(size = 4, alpha = 0.8) +
  geom_text(aes(label = PatientID), vjust = 2, size = 3, show.legend = FALSE) +
  theme_bw() +
  
  # Assign complementary light/dark shades so the diseases group visually
  scale_color_manual(
    name = "Disease & Sex",
    values = c(
      "ASD_Female" = "orange", 
      "ASD_Male" = "firebrick", 
      "TOF_Female" = "deepskyblue", 
      "TOF_Male" = "darkblue"
    )
  ) +
  
  labs(
    title = "PCA of Filtered Expression Data",
    subtitle = "Single Panel: Disease/Sex (Color) and Time (Shape)",
    x = paste0("PC1 (", round(summary(pca_res)$importance[2,1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_res)$importance[2,2] * 100, 1), "%)")
  )
```

## Batch Effect Assessment
Physical scan dates were extracted to assess temporal batch effects.
```{r batch-effect-check}
table(Group = metadata$SampleGroup, Date = metadata$ScanDate)

pca_plot_data$ScanDate <- metadata$ScanDate
pca_plot_data$ScanYear <- metadata$ScanYear

ggplot(pca_plot_data, aes(x = PC1, y = PC2, color = ScanDate, shape = Time)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_bw() +
  labs(
    title = "PCA Checking for Batch Effects (Scan Date)",
    x = paste0("PC1 (", round(summary(pca_res)$importance[2,1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_res)$importance[2,2] * 100, 1), "%)")
  )

year_colors <- c("2013" = "dodgerblue", "2014" = "firebrick")

ggplot(pca_plot_data, aes(x = PC1, y = PC2, color = ScanYear, shape = Time)) +
  geom_point(size = 4, alpha = 0.8) +
  theme_bw() +
  scale_color_manual(values = year_colors) +
  labs(
    title = "PCA Checking for Batch Effects (Scan Year)",
    x = paste0("PC1 (", round(summary(pca_res)$importance[2,1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_res)$importance[2,2] * 100, 1), "%)")
  )
```
## Batch Effect Correction
```{r batch-effect-correction}
metadata$group <- factor(paste(metadata$Disease, metadata$Time, metadata$Sex, sep = "_"))

# Combat correction for ScanYear
mod_combat <- model.matrix(~ group, data = metadata)

expression_matrix_filtered <- ComBat(
  dat = expression_matrix_filtered, 
  batch = metadata$ScanYear, 
  mod = mod_combat, 
  par.prior = TRUE
)

# SVA for hidden batches
mod_sva  <- model.matrix(~ group, data = metadata)
mod0_sva <- model.matrix(~ 1, data = metadata)          

# Estimate the number of hidden surrogate variables on the COMBAT-SCRUBBED data
n_sv_conservative <- num.sv(expression_matrix_filtered, mod_sva, method = "be")
message("The conservative BE method recommends: ", n_sv_conservative, " additional hidden batches.")

# Calculate the Surrogate Variables
svobj <- sva(expression_matrix_filtered, mod_sva, mod0_sva)

# Re-calculate PCA on the newly scrubbed ComBat matrix
pca_res_post <- prcomp(t(expression_matrix_filtered), scale. = TRUE)
pca_df_post <- as.data.frame(pca_res_post$x)
pca_plot_data_post <- cbind(pca_df_post, metadata)

# Define the batch colors again
year_colors <- c("2013" = "dodgerblue", "2014" = "firebrick")

# Generate the plot
ggplot(pca_plot_data_post, aes(x = PC1, y = PC2, color = ScanYear, shape = Time)) +
  geom_point(size = 4, alpha = 0.8) +
  theme_bw() +
  scale_color_manual(values = year_colors) +
  labs(
    title = "Post-ComBat PCA: Batch Effect Assessment",
    x = paste0("PC1 (", round(summary(pca_res_post)$importance[2,1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_res_post)$importance[2,2] * 100, 1), "%)")
  )

# Combined plot
pca_res_final <- prcomp(t(expression_matrix_filtered), scale. = TRUE)
pca_df_final <- as.data.frame(pca_res_final$x)

# Bind with metadata
pca_plot_data_final <- cbind(pca_df_final, metadata)
pca_res.var_final <- pca_res_final$sdev^2
pca_res.var.per_final <- round(pca_res.var_final/sum(pca_res.var_final)*100, 1)

barplot(pca_res.var.per_final, 
        main = "Post-ComBat Scree Plot", 
        xlab = "Principal Component", 
        ylab = "Percent Variation",
        col = "steelblue")

pca_plot_data_final$Disease_Sex <- paste(pca_plot_data_final$Disease, pca_plot_data_final$Sex, sep = "_")

ggplot(pca_plot_data_final, aes(x = PC1, y = PC2, color = Disease_Sex, shape = Time)) +
  geom_point(size = 4, alpha = 0.8) +
  theme_bw() +
  
  scale_color_manual(
    name = "Disease & Sex",
    values = c(
      "ASD_Female" = "orange", 
      "ASD_Male" = "firebrick", 
      "TOF_Female" = "deepskyblue", 
      "TOF_Male" = "darkblue"
    )
  ) +
  
  labs(
    title = "Post-ComBat PCA: Disease, Sex, and CPB Response",
    x = paste0("PC1 (", round(summary(pca_res_final)$importance[2,1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_res_final)$importance[2,2] * 100, 1), "%)")
  )

```

## Limma DGE
``` {r limma-model}
# Primary grouping factor, ex: TOF_Post_Male
metadata$group <- factor(paste(metadata$Disease, metadata$Time, metadata$Sex, sep = "_"))

design <- model.matrix(~ 0 + group, data = metadata)
colnames(design) <- gsub("group", "", colnames(design))

# Append the SVs to the design matrix
if(svobj$n.sv > 0) {
  design <- cbind(design, svobj$sv)
  colnames(design)[(ncol(design) - svobj$n.sv + 1):ncol(design)] <- paste0("SV", 1:svobj$n.sv)
  message(paste("SVA found and appended", svobj$n.sv, "hidden batch effect(s) to the design matrix."))
} else {
  message("SVA found 0 hidden batches.")
}

aw <- arrayWeights(expression_matrix_filtered, design)

# Plot arrayWeights
names(aw) <- colnames(expression_matrix_filtered)
par(mar = c(12, 4, 4, 2)) 
barplot(aw, 
        las = 2, 
        names.arg = names(aw), # FIX: Force the plot to use the names
        col = ifelse(grepl("ASD_Pre_Female_Pt4", names(aw)), "firebrick", "steelblue"),
        main = "Limma Array Weights: Sample Quality",
        ylab = "Relative Weight (1.0 is average)",
        cex.names = 0.8) 
abline(h = 1, col = "red", lty = 2)
par(mar = c(5, 4, 4, 2) + 0.1)

# duplicateCorrelation estimates the internal consistency of each child's samples
corfit <- duplicateCorrelation(expression_matrix_filtered, design, block = metadata$PatientID)

fit <- lmFit(expression_matrix_filtered, design, block = metadata$PatientID, correlation = corfit$consensus, weights = aw)

# Create contrasts
contrastMatrix <- makeContrasts(
  TOF_Male_Response = TOF_Post_Male - TOF_Pre_Male,
  TOF_Female_Response = TOF_Post_Female - TOF_Pre_Female,
  ASD_Male_Response = ASD_Post_Male - ASD_Pre_Male,
  ASD_Female_Response = ASD_Post_Female - ASD_Pre_Female,
  
  # Sex diff
  TOF_Sex_Diff = (TOF_Post_Female - TOF_Pre_Female) - (TOF_Post_Male - TOF_Pre_Male),
  ASD_Sex_Diff = (ASD_Post_Female - ASD_Pre_Female) - (ASD_Post_Male - ASD_Pre_Male),
  
  levels = design
)

fit2 <- contrasts.fit(fit, contrastMatrix)
fit2 <- eBayes(fit2, robust = TRUE)

# View Summaries
resultsSummary <- decideTests(fit2)
print(summary(resultsSummary))
vennDiagram(resultsSummary[, c("TOF_Male_Response", "TOF_Female_Response", "ASD_Male_Response", "ASD_Female_Response")])
```
## Variance Partitioning
This identifies which variables (Disease, Sex, Time) drive the most expression changes
```{r variance-partition, message = FALSE}
# setup parallel processing to speed up calculations
param <- SnowParam(workers = 4, progressbar = TRUE)
register(param)

# Include ScanYear to verify ComBat successfully scrubbed the batch effect
info <- metadata %>%
  dplyr::select(Disease, Sex, Time, ScanYear, PatientID) %>%
  as.data.frame()
rownames(info) <- colnames(expression_matrix_filtered)

if(exists("svobj") && svobj$n.sv > 0) {
  # Add each SV as a new column in the info dataframe
  for (j in 1:svobj$n.sv) {
    info[[paste0("SV", j)]] <- svobj$sv[, j]
  }
  
  # Build the formula string with continuous SVs as fixed effects (no '1|')
  sv_terms <- paste0("SV", 1:svobj$n.sv, collapse = " + ")
  form_string <- paste("~ (1|Disease) + (1|Sex) + (1|Time) + (1|ScanYear) +", sv_terms)
  
} else {
  # Fallback formula if SVA found 0 hidden batches
  form_string <- "~ (1|Disease) + (1|Sex) + (1|Time) + (1|ScanYear)"
}

form <- as.formula(form_string)

varPart <- fitExtractVarPartModel(expression_matrix_filtered, form, info, weights = aw)
```

## Plot Variance Partition Results
``` {r plot-vp} 
vp <- sortCols(varPart)

vp_df <- as.data.frame(vp)
vp_df$Gene_Symbol <- hgnc_symbols_filtered[match(rownames(vp_df), names(hgnc_symbols_filtered))]

vp_final <- vp_df %>%
  dplyr::select(Gene_Symbol, everything()) %>%
  dplyr::arrange(desc(Time))

vp_for_plot <- vp_final[1:20, ]
unique_labels <- paste0(vp_for_plot$Gene_Symbol, " (", rownames(vp_for_plot), ")")
rownames(vp_for_plot) <- unique_labels
vp_for_plot$Gene_Symbol <- NULL # Remove the column so only numbers remain

plotPercentBars(as.matrix(vp_for_plot))
plotVarPart(vp, main = "Variance Partition: CPB Response Drivers")

message("Top genes driven by CPB Response (Pre vs Post):")
head(vp_final[order(vp_final$Time, decreasing = TRUE), ])
```

## Model Pruning: Selecting Significant Surrogate Variables
SVA identified 10 potential surrogate variables. However based on the variance partition analysis, only SV1 contributes meaningfully (>10% median variance). To preserve statistical power (degrees of freedom) in our cohort, we will prune the model to include only SV1 as a covariate, discarding the negligible artifacts.

```{r limma-model-pruned}
design <- model.matrix(~ 0 + group, data = metadata)
colnames(design) <- gsub("group", "", colnames(design))

# Append only SV1 to the design matrix
design <- cbind(design, SV1 = svobj$sv[, 1])
message("Final Design Matrix Columns: ", paste(colnames(design), collapse = ", "))

# Recalculate weights and correlation using the pruned design
aw <- arrayWeights(expression_matrix_filtered, design)

# Plot arrayWeights
names(aw) <- colnames(expression_matrix_filtered)
par(mar = c(12, 4, 4, 2)) 
barplot(aw, 
        las = 2, 
        names.arg = names(aw), # FIX: Force the plot to use the names
        col = ifelse(grepl("ASD_Pre_Female_Pt4", names(aw)), "firebrick", "steelblue"),
        main = "Limma Array Weights: Sample Quality",
        ylab = "Relative Weight (1.0 is average)",
        cex.names = 0.8) 
abline(h = 1, col = "red", lty = 2)
par(mar = c(5, 4, 4, 2) + 0.1)

corfit <- duplicateCorrelation(expression_matrix_filtered, design, block = metadata$PatientID)

fit <- lmFit(expression_matrix_filtered, design, 
             block = metadata$PatientID, 
             correlation = corfit$consensus, 
             weights = aw)

contrastMatrix <- makeContrasts(
  TOF_Male_Response = TOF_Post_Male - TOF_Pre_Male,
  TOF_Female_Response = TOF_Post_Female - TOF_Pre_Female,
  ASD_Male_Response = ASD_Post_Male - ASD_Pre_Male,
  ASD_Female_Response = ASD_Post_Female - ASD_Pre_Female,
  
  # Sex diff
  TOF_Sex_Diff = (TOF_Post_Female - TOF_Pre_Female) - (TOF_Post_Male - TOF_Pre_Male),
  ASD_Sex_Diff = (ASD_Post_Female - ASD_Pre_Female) - (ASD_Post_Male - ASD_Pre_Male),
  
  # Disease diff
  Disease_Diff = (((TOF_Post_Male - TOF_Pre_Male) + (TOF_Post_Female - TOF_Pre_Female))/2 - ((ASD_Post_Male - ASD_Pre_Male) + (ASD_Post_Female - ASD_Pre_Female))/2),
  
  # Baseline diff by sex
  TOF_pre_surg = (TOF_Pre_Female - TOF_Pre_Male),
  ASD_pre_surg = (ASD_Pre_Female - ASD_Pre_Male),
  
  # Post surgery diff by sex
  TOF_post_surg = (TOF_Post_Female - TOF_Post_Male),
  ASD_post_surg = (ASD_Post_Female - ASD_Post_Male),
  
  levels = design
)

fit2 <- contrasts.fit(fit, contrastMatrix)
fit2 <- eBayes(fit2, robust = TRUE)

# View Summaries
resultsSummary <- decideTests(fit2)
print(summary(resultsSummary))
vennDiagram(resultsSummary[, c("TOF_Male_Response", "TOF_Female_Response", "ASD_Male_Response", "ASD_Female_Response")])
```

## Visualizing removal of SV1
```{r visualize-cleaned-data}
design_keep <- model.matrix(~ 0 + group, data = metadata)

# Add the SV to covariates, not batch
expression_matrix_clean <- removeBatchEffect(
  expression_matrix_filtered, 
  covariates = svobj$sv[, 1],
  design = design_keep
)

# Run PCA
pca_clean <- prcomp(t(expression_matrix_clean), scale. = TRUE)
pca_df <- as.data.frame(pca_clean$x)

pca_plot_data <- cbind(pca_df, metadata) %>%
  mutate(Disease_Sex = paste(Disease, Sex, sep = "_"))

ggplot(pca_plot_data, aes(x = PC1, y = PC2, color = Disease_Sex, shape = Time)) +
  geom_point(size = 4, alpha = 0.7) + 
  theme_bw() +
  scale_color_manual(
    values = c(
      "ASD_Female" = "orange", 
      "ASD_Male" = "firebrick", 
      "TOF_Female" = "deepskyblue", 
      "TOF_Male" = "darkblue"
    )
  ) +
  labs(
    title = "PCA Post SV1 Removal",
    x = paste0("PC1 (", round(summary(pca_clean)$importance[2,1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_clean)$importance[2,2] * 100, 1), "%)")
  )
```
## Re-Run Variance Partioning
``` {r re-run-var-partioning}
varPart_validation <- fitExtractVarPartModel(
  expression_matrix_clean, 
  form, 
  info, 
  weights = aw
)
```
``` {r plot-var-part-plot}
vp_valid <- sortCols(varPart_validation)
plotVarPart(vp_valid, main = "Variance Partition After SV1 Removal")
```
## Post-Analysis and Result Generation
```{r post-analysis-pruned}
output_dir <- "results"
dir.create(output_dir, showWarnings = FALSE)
hgnc_complete_set_df <- read_tsv("hgnc_complete_set.txt") %>%
  dplyr::select(HGNC_Symbol = symbol)

contrasts_list <- colnames(contrastMatrix)
for (contrast_name in contrasts_list) {
  res_df <- topTable(fit2, coef = contrast_name, number = Inf, adjust.method = "BH")
  
  # Add HGNC gene symbol
  res_df_with_hgnc <- res_df %>%
    as_tibble(rownames = "Original_Probe_ID") %>% 
    mutate(HGNC_Symbol = hgnc_symbols_filtered[match(Original_Probe_ID, names(hgnc_symbols_filtered))])
  
  # Remove rows where HGNC_Symbol is NA
  res_df_with_hgnc <- res_df_with_hgnc %>%
    filter(!is.na(HGNC_Symbol))
  
  # Perform inner_join with the hgnc_complete_set data
  joined_results <- res_df_with_hgnc %>%
    inner_join(hgnc_complete_set_df %>% dplyr::select(HGNC_Symbol), by = "HGNC_Symbol")
  
  final_results_table <- joined_results %>%
    dplyr::arrange(adj.P.Val) %>% 
    dplyr::distinct(HGNC_Symbol, .keep_all = TRUE)
  
  final_results_table <- final_results_table %>%
    dplyr::select(HGNC_Symbol, logFC, P.Value, adj.P.Val)
  
  final_results_table <- final_results_table[!is.na(final_results_table$adj.P.Val), ]
  
  file_name <- paste0(output_dir, "/dge_results_", contrast_name, "_final.csv")
  write_csv(final_results_table, file_name)
  
  message(paste("Saved:", file_name))
}
```
